External priors for the next generation of CMB experiments.

### 2016-02-22 Notes/response.txt

- The baseline dataset in all the results is using CMB+BAO which means
already CMB + external information. This is confusing, we are not
improving on CMB then but rather on a combination of CMB and some
galaxy data.

The referee is right; we were not clear that CMB+BAO includes current BAO constraints, from e.g., BOSS.
It doesn't make sense to us to exclude things we currently know, so we think it important to include these constraints as our baseline. For example that will strongly bias the foracasted performance of CMB S4 when considered without future BAO experiments like DESI.

- I cannot understand Figures 1,2,3,4. If the far right is the
expected CMB S4+Planck-Pol+BAO15 then the lines starting point should
correspond to prior/sigma = 1. How is it possible than adding a prior
on a parameter 3 times bigger than the sigma from CMB
S4+Planck-Pol+BAO15 (e.g., on Ho) gives a better constrain? If I think
about it in the likelihood space I am multiplying a distribution with
another one having 3 times its width and I do not expect to have any
impact on the combined result. Also, I do not understand why we should
consider adding a prior which is bigger than the baseline sigma from
CMB S4+Planck-Pol+BAO. What is shown on the right axis?

We agree that this was unnecessarily complicated so have changed all the figures to absolute priors on the parameters.
We also explicitly include the the fractional error on right axis in addition to the absolute error on the left axis.
The referee's question still remains: why does a 3% prior on H_0, say, help if the data itself constrain it to 1%. We stand by our result that it does help. Imagine a 1-D likelihood with one experiment giving a 1% constraint and another [here called "external"] a similar 1% constraint. When combining the 2, the constraint will be tighter, with the new error given by the inverse square root of the sum of the two inverse variances: [1/\sigma_1^2 + 1/\sigma_2^2]^{-1/2}. In the simple example above the external prior is equal to the internal constraint, so the total constraint is reduced by \sqrt{2}. In the example given by the referee [external constraint 3 times looser], there will still be improvement, albeit only by a factor of \sqrt{1 + 1/9}, ~5%. In multi-dimensional space, things are not that simple, but the plots show how things improve. To come back to our first response, we believe this information is extremely important for future planning. While it is true that people have studied degeneracies before, their impact has not been presented in this way. The very fact that the plots are not familiar indicates that we are introducing a new way, we think a productive way, of thinking about planning. We have added a paragraph in the text that attempts to explain this point more clearly and -- as noted -- have changed the x-axis on all plots to make things less confusing.

@@

- In the introduction: Ref. 4 is WMAP 1year results, in this context
the general WMAP experiment paper Bennett et al. 2003 or the final
mission results Bennet et al. 2012 would be more appropriate. Same
thing for Planck, cite the Blue Book and/or the 2015 Overview papers.
Ref. 5 is Maxima and Ref. 6 is Boomerang, they are not satellites and
are not mentioned in the text.

Thanks, Fixed adding Bennett et al. 2003 and Blue book

- ACT and SPT did not follow DASI, they have measured temperature with
ACTPol and SPTPol going for polarization. The first two should go into
the above temperature discussion with references to latest results or
general papers describing the experiments, and the last two should
contribute to the polarization discussion.

We were not focused so much on the experiments but the achievements that feed into the projections in this paper. First, there was detection of polarization; then measurement of small scale anisotropies that yield the damping tail and the lensing maps. Reworded taking out explicit mention of the experiments to be clearer.

@@

- Section IIIA “power power spectrum is suppressed” -> “power spectrum
is suppressed”

Thanks, done.

- Section IIIA, what is the meaning of “since the active neutrino
number density are known in the standard model”?

We meant that the number density of active neutrinos is a robust prediction emerging from the Standard Model of particle physics. Capitalized to make a little clearer

@@

- Fig. 5 caption, there is no red line in the plot
We fixed it.

- Last part of Section IV, cite LSST.

Done (http://arxiv.org/abs/1211.0310)

- Last part of Section IV, what do the authors mean with surveys
beyond DESI and LSST? LSST is beyond DESI and I am not aware of
post-LSST surveys. Also, galaxy surveys will not help in constraining
n_s.

Part of the point here is to motivate surveys that extract more of the information on the sky than will DESI and LSST. We agree with the referee that such surveys [galaxy or otherwise] are not currently planned, but we think the arguments here can help motivate people to start planning them. Already, several of the plots in this paper were shown at the NRAO 2020 Futures meeting [https://science.nrao.edu/science/meetings/2015/2020futures/program]

---------------------

### 2016-02-22 Notes/CMBS4_priors_paper.tex
working on response

\section{Introduction}\label{sec:intro}
Since their earliest incarnations, Cosmic Microwave Background (CMB) experiments have been crucial in furthering our understanding of the Universe. They will maintain their role in the future, thanks to the unprecedented low level of noise and high resolution expected in the planned Stage IV (S4) experiment~\cite{2013arXiv1309.5383A}.
The recent past is reassuring. For example, every new generation of satellite experiments improved the level of sensitivity by almost a factor of ten compared to its predecessor, from the first generation instrument COBE to WMAP all the way to the current state of the art represented by Planck \cite{fixsen:1996,BlueBook2005, bennett:2003}. 
These led us from the first detection of anisotropy in the CMB temperature with COBE to a cosmic variance limited measurement of several acoustic peaks with Planck. With the improved sensitivity we have extended our understanding of the Universe so that we now have solid evidence for a flat geometry and a well tested $\Lambda$CDM model  with percent level constraints on its parameters.
The same improvement has characterized ground-based CMB experiments, where, e.g., the first detection of polarization anisotropy by DASI \cite{2002ApJ...568...38H} has been followed by a series of precise measurements of the damping tail and maps of the projected gravitational potential by, e.g., SPT~\cite{2011ApJ...743...28K} and ACT~\cite{2011ApJ...739...52D}.
The planned CMB-S4 experiment will potentially measure the E-mode polarization with cosmic variance limited precision together with an order of magnitude improvement in B-mode measurement and lensing reconstruction. As has happened in the past, this new sensitivity together with the progress in the measurement of other cosmological probes will improve our understanding of dark matter, inflation, dark energy, neutrinos, and other  Beyond the Standard Model physics.

The cosmological dependence on neutrinos is two-fold: the number of relativistic species $N_{\rm eff}$ in the early phase of the Universe affects the damping tail of the CMB, and sum of the neutrino masses $\sum m_\nu$ affects the late-time growth of cosmic structure.
The sensitivity to $N_{\rm eff}$ is due to its affects on the expansion rate $H(z)$ because relativistic species, like neutrinos, are the main drivers of the cosmic expansion in the early Universe. This rate can be powerfully tested using the CMB, by carefully comparing the sound horizon scale, obtained from the CMB peaks positions, and the Silk damping scale (see \cite{2013arXiv1309.5383A}, \cite{2013PhRvD..87h3008H} and references therein). 
The future experiments, with their high resolution, will probe deep into the damping tale of the CMB power spectrum and will be sensitive to small variations from the canonical, 3-active neutrino prediction of $N_{\rm eff}=3.046$. On the other hand, the total mass of neutrinos, has a modest effect on the CMB because, for the range of masses allowed by recent constraints ($\sum m_{\nu}<230$ meV from \cite{planck-collaboration:2014}), neutrinos are still relativistic at the last scattering surface. However, massive neutrinos alter the growth of the large scale structure responsible for lensing of the CMB photons. Different neutrino masses consequently lead to different CMB lensing spectra. CMB-S4 will measure small scale temperature and polarization anisotropies with low noise,  improving by at least one order of magnitude the current lensing reconstruction. This will turn into a precise constraints of the sum of neutrino masses with a possible hint of the hierarchies of the individual masses.

@@

In the cases of both neutrinos and dark energy then, an important lingering question is: What external information can be used to improve projected constraints by breaking degeneracies?
Here we study the dependance of projected constraints from CMB on the external priors assumed. For all the parameters of the standard $\Lambda$CDM with massive neutrinos ($\Omega_ch^2, \Omega_bh^2, A_s, n_s, \tau, H_0, \sum m_\nu$) together with extensions ($N_{\rm eff}$ and $w$) to the neutrino and dark energy sectors, we quantify the CMB constraints as a function of external priors. This extends the work of \citet{wu:2014}, which worked with a few fixed external priors, by quantifying the extent to which external information will improve the constraining power of a CMB-S4 experiment.

@@

Small scale structure formation is suppressed if fast-moving neutrinos comprise a significant part of the matter budget. 
Indeed, below the free-streaming scale, the matter power spectrum is suppressed in the presence of
massive neutrinos by a factor $\Delta P/P\simeq -8f_{\nu}$ , where $f_{\nu} = \Omega_{\nu} / \Omega_{m}$ is the contribution of neutrinos to the total matter density.
This effect is probed by CMB lensing and, since the active neutrino number densities are known in the standard model, can be directly transformed into a constraint on the sum of the neutrino masses $\sum m_\nu$. This is particularly exciting because we now know that neutrinos are massive, with a lower limit $\sum m_\nu>50$ meV that emerges from oscillation experiments (see \cite{bellini:2013} and references therein).

@@

\reffig{prior_omeganuh2-h} and \reffig{prior_omeganuh2-omch2} show the projected errors on $\sum m_\nu$ from CMB-S4 (+Planck+current BAO) as a function of external priors on $H_{0}$ and $\Omega_{c}h^{2}$. 
Without any external priors (far right in figure), upcoming CMB experiments are poised to obtain 1-sigma limits on $\sum m_\nu$ of order 25 meV, corresponding to a 2-sigma detection even in the worst case scenario. Adding in external priors helps this significantly though, as discussed, for example, in \cite{di-valentino:2015,2013arXiv1309.5383A,pan:2015a,allison:2015}. 
There, the added prior was ``DESI BAO'', representing a measurement of distances as a function of redshift from the baryon acoustic oscillation feature probed by the planned experiment DESI \cite{levi:2013}. This is represented by the blue band in our figures. In addition to showing the implications of introducing this fixed external constraint, \reffig{prior_omeganuh2-h} and \reffig{prior_omeganuh2-omch2} generalize the extent to which external information will improve the neutrino mass constraints. For example, in \reffig{prior_omeganuh2-h} the red curve shows the improvement as additional information on the Hubble constant is applied. We can see, for example that, imposing {\it external} information with at the level of $\sigma_{H_0}/H_0=0.6\%$ would improve the neutrino mass constraint to 17 meV. Similarly, in \reffig{prior_omeganuh2-omch2} adding to CMB S4 + BAO15 an external constraint on the dark matter density equal to its internal constraint of $0.4\%$ improves the projected neutrino mass constraint to 20 meV. It is not surprising that this is close to the constraint obtained with DESI because the BAO distances in this model depends most sensitively on $\Omega_ch^2$ and $H_{0}$.

>>>>

\section{Introduction}\label{sec:intro}
Since their earliest incarnations, Cosmic Microwave Background (CMB) experiments have been crucial in furthering our understanding of the Universe. They will maintain their role in the future, thanks to the unprecedented low level of noise and high resolution expected in the planned Stage IV (S4) experiment~\cite{2013arXiv1309.5383A}.
The recent past is reassuring. For example, every new generation of satellite experiments improved the level of sensitivity by almost a factor of ten compared to its predecessor, from the first generation instrument COBE to WMAP all the way to the current state of the art represented by Planck \cite{fixsen:1996,BlueBook2005, bennett:2003}. 
These led us from the first detection of anisotropy in the CMB temperature with COBE to a cosmic variance limited measurement of several acoustic peaks with Planck. With the improved sensitivity we have extended our understanding of the Universe so that we now have solid evidence for a flat geometry and a well tested $\Lambda$CDM model  with percent level constraints on its parameters.
Similar progress has characterized ground-based CMB experiments, where the first detection of polarization anisotropy \cite{2002ApJ...568...38H} has been followed by a series of precise measurements of the damping tail~\cite{2011ApJ...739...52D,2011ApJ...743...28K} and of the projected gravitational potential~\cite{2011PhRvL.107b1301D,2012ApJ...756..142V}.

The planned CMB-S4 experiment will potentially measure the E-mode polarization with cosmic variance limited precision together with an order of magnitude improvement in B-mode measurement and lensing reconstruction. As has happened in the past, this new sensitivity together with the progress in the measurement of other cosmological probes will improve our understanding of dark matter, inflation, dark energy, neutrinos, and other  Beyond the Standard Model physics.

The cosmological dependence on neutrinos is two-fold: the number of relativistic species $N_{\rm eff}$ in the early phase of the Universe affects the damping tail of the CMB, and sum of the neutrino masses $\sum m_\nu$ affects the late-time growth of cosmic structure.
The sensitivity to $N_{\rm eff}$ is due to its effects on the expansion rate $H(z)$ because relativistic species, like neutrinos, are the main drivers of the cosmic expansion in the early Universe. This rate can be powerfully tested using the CMB, by carefully comparing the sound horizon scale, obtained from the CMB peaks positions, and the Silk damping scale (see \cite{2013arXiv1309.5383A}, \cite{2013PhRvD..87h3008H} and references therein). 
The future experiments, with their high resolution, will probe deep into the damping tale of the CMB power spectrum and will be sensitive to small variations from the canonical, 3-active neutrino prediction of $N_{\rm eff}=3.046$. On the other hand, the total mass of neutrinos, has a modest effect on the CMB because, for the range of masses allowed by recent constraints ($\sum m_{\nu}<230$ meV from \cite{planck-collaboration:2014}), neutrinos are still relativistic at the last scattering surface. However, massive neutrinos alter the growth of the large scale structure responsible for lensing of the CMB photons. Different neutrino masses consequently lead to different CMB lensing spectra. CMB-S4 will measure small scale temperature and polarization anisotropies with low noise,  improving by at least one order of magnitude the current lensing reconstruction. This will turn into a precise constraints of the sum of neutrino masses with a possible hint of the hierarchies of the individual masses.

@@

In the cases of both neutrinos and dark energy then, an important lingering question is: What external information can be used to improve projected constraints by breaking degeneracies?
Here we study the dependance of projected constraints from CMB on the external priors assumed. The canonical set of parameters is: cold dark matter density $\Omega_ch^2$; baryon density $\Omega_bh^2$; amplitude of primordial perturbations $A_s$; slope of primordial spectrum  $n_s$; optical depth to the last scattering surface $\tau$; Hubble constant $H_0$; and sum of the neutrino masses $\sum m_\nu$.  Together with extensions (the number of relativistic degrees of freedom $N_{\rm eff}$ and dark energy equation of state $w$) to the neutrino and dark energy sectors, we quantify the CMB constraints as a function of external priors. This extends the work of \citet{wu:2014}, which worked with a few fixed external priors, by quantifying the extent to which external information will improve the constraining power of a CMB-S4 experiment.

@@

Small scale structure formation is suppressed if fast-moving neutrinos comprise a significant part of the matter budget. 
Indeed, below the free-streaming scale, the matter power spectrum is suppressed in the presence of
massive neutrinos by a factor $\Delta P/P\simeq -8f_{\nu}$ , where $f_{\nu} = \Omega_{\nu} / \Omega_{m}$ is the contribution of neutrinos to the total matter density.
This effect is probed by CMB lensing and, since the active neutrino number densities are known in the Standard Model, can be directly transformed into a constraint on the sum of the neutrino masses $\sum m_\nu$. This is particularly exciting because we now know that neutrinos are massive, with a lower limit $\sum m_\nu>50$ meV that emerges from oscillation experiments (see \cite{bellini:2013} and references therein).

@@

\reffig{prior_omeganuh2-h} shows the projected errors on $\sum m_\nu$ from CMB-S4 (+Planck+current BAO) as a function of external priors on $H_{0}, \Omega_{c}h^{2}, n_s$ and $A_s$. 
Without any external priors (far right in figure), upcoming CMB experiments are poised to obtain 1-sigma limits on $\sum m_\nu$ of order 25 meV, corresponding to a 2-sigma detection even in the worst case scenario. Adding in external priors helps this significantly though, as discussed, for example, in \cite{di-valentino:2015,2013arXiv1309.5383A,pan:2015a,allison:2015}. 
There, the added prior was ``DESI BAO'', representing a measurement of distances as a function of redshift from the baryon acoustic oscillation feature probed by the planned experiment DESI \cite{levi:2013}. This is represented by the blue band in our figures. In addition to showing the implications of introducing this fixed external constraint, \reffig{prior_omeganuh2-h} generalizes the extent to which external information will improve the neutrino mass constraints. For example, in \reffig{prior_omeganuh2-h} the red solid curve shows the improvement as additional information on the Hubble constant is applied. We can see, for example that, imposing {\it external} information at the level of $\sigma_{H_0}/H_0=0.5\%$ would improve the neutrino mass constraint to 17 meV. Similarly, in \reffig{prior_omeganuh2-omch2} adding to CMB S4 + BAO15 an external constraint on the dark matter density equal to its internal constraint of $0.4\%$ improves the projected neutrino mass constraint to 20 meV. It is not surprising that this is close to the constraint obtained with DESI because the BAO distances in this model depends most sensitively on $\Omega_ch^2$ and $H_{0}$.
